A stable, distributed code for cue value in mouse cortex during reward learning

Version 1

Abstract

Summary

The ability to associate reward-predicting stimuli with adaptive behavior is frequently attributed to the prefrontal cortex, but the stimulus-specificity, spatial distribution, and stability of pre-frontal cue-reward associations are unresolved. We trained headfixed mice on an olfactory Pavlovian conditioning task and measured the coding properties of individual neurons across space (prefrontal, olfactory, and motor cortices) and time (multiple days). Neurons encoding cues or licks were most common in olfactory and motor cortex, respectively. By quantifying the responses of cue-encoding neurons to six cues with varying probabilities of reward, we unexpectedly found value coding, including coding of trial-by-trial reward history, in all regions we sampled. We further found that prefrontal cue and lick codes were preserved across days. Our results demonstrate that individual prefrontal neurons stably encode components of cue-reward learning within a larger spatial gradient of coding properties.

Introduction

Association of environmental stimuli with rewards and the subsequent orchestration of value-guided reward-seeking behavior are crucial functions of the nervous system linked to the pre-frontal cortex (PFC) (Klein-Flügge et al., 2022; Miller and Cohen, 2001). PFC is heterogeneous, with many studies noting subregional differences in both neural coding of (Hunt et al., 2018; Kennerley et al., 2009; Sul et al., 2010; Wang et al., 2020a) and functional impact on (Buckley et al., 2009; Dalley et al., 2004; Kesner and Churchwell, 2011; Rudebeck et al., 2008) value-based reward seeking in primates and rodents. Further, functional manipulations of PFC subregions exhibiting robust value signals do not always cause a discernible impact on reward-guided behavior (Chudasama and Robbins, 2003; Dalton et al., 2016; St. Onge and Floresco, 2010; Verharen et al., 2020; Wang et al., 2020a), encouraging investigation of differences between value signals across PFC. Within individual PFC subregions, multiple studies have observed evolving neural representations across time, calling into question the stability PFC signaling (Hyman et al., 2012; Malagon-Vina et al., 2018). A systematic comparison of coding properties across rodent PFC and related motor and sensory regions, as well as across days and stimulus sets, is necessary to provide a full context for the contributions of PFC subregions to reward processing.

Identifying neural signals for value requires a number of considerations. One issue is that other task features can vary either meaningfully or spuriously with value. In particular, action coding is difficult to parse from value signaling, given the high correlations between behavior and task events (Musall et al., 2019; Zagha et al., 2022) and widespread neural coding of reward-seeking actions (Steinmetz et al., 2019). Additionally, without a sufficiently rich value axis, it is possible to misidentify neurons as ‘value’ coding even though they do not generalize to valuations in other contexts (Hayden and Niv, 2021; Stalnaker et al., 2015; Zhou et al., 2021). Because reports of value have come from different experiments across different species, it is difficult to compare the presence of value signaling even across regions within prefrontal cortex (Hayden and Niv, 2021; Hunt et al., 2018; Kennerley et al., 2009; Namboodiri et al., 2019; Otis et al., 2017; Stalnaker et al., 2015; Sul et al., 2010; Wang et al., 2020a; Zhou et al., 2021).

In this work, we sought to address the existing ambiguity in the distribution and stability of value signaling. We implemented an olfactory Pavlovian conditioning task that permitted identification of value correlates within the domain of reward probability across two separate stimulus sets. With acute in vivo electrophysiology recordings, we were able to assess coding of this task across 11 brain regions, including five PFC subregions, as well as olfactory and motor cortex, in a single group of mice, permitting a well-controlled comparison of coding patterns across a large group of task-relevant regions in the same subjects. Unexpectedly, in contrast to the graded cue and lick coding across these regions, there was a similar proportion of neurons encoding cue value in all regions. A subset of value coding neurons were sensitive to trial-by-trial fluctuations in value according to reward history. To assess coding stability, we performed 2-photon imaging of neurons in PFC for multiple days and determined that the cue and lick codes we identified were stable over time. Our data demonstrate universality and stability of cue-reward coding in mouse cortex.

Results

Distributed neural activity during an olfactory Pavlovian conditioning task

We trained mice on an olfactory Pavlovian conditioning task with three cue (conditioned stimulus) types that predicted reward on 100% (‘CS+’), 50% (‘CS50’), or 0% (‘CS−’) of trials (Fig. 1A). Each mouse learned two odor sets, trained and imaged on separate days and then, for electrophysiology experiments, presented in six alternating blocks of 51 trials during recording sessions (Fig. 1B). Mice developed anticipatory licking (Fig. 1C-D), and the rate of this licking correlated with reward probability (Fig. S1), indicating that subjects successfully learned the meaning of all six odors.

Electrophysiology and imaging during olfactory Pavlovian conditioning.

(A) Trial structure in Pavlovian conditioning task.

(B) Timeline for mouse training.

(C) Mean (+/− standard error of the mean (SEM)) lick rate across mice (n = 5) on each trial type for each odor set during electrophysiology sessions. CS50(r) and CS50(u) are rewarded and unrewarded trials, respectively. Inset: mean anticipatory licks (change from baseline) for the CS+ and CS50 cues for every session, color-coded by mouse. F (1, 66) = 36.6 for a main effect of cue in a two-way ANOVA including an effect of subject.

(D) Same as (C), for Day 3 imaging sessions (n = 5 mice). t(4) = −5.4 for a t-test comparing anticipatory licks on CS+ and CS50 trials.

(E) Neuropixels probe tracks labeled with fluorescent dye (red) in cleared brain (autofluorescence, green). AP, anterior/posterior; ML, medial/lateral; DV, dorsal/ventral. Allen CCF regions delineated in gray. Outline of prelimbic area in purple.

(F) Reconstructed recording sites from all tracked probe insertions (n = 44 insertions, n = 5 mice), colored by mouse.

(G) Sample histology image of lens placement. Visualization includes DAPI (blue) and GCaMP (green) signal with lines indicating cortical regions from Allen Mouse Brain Common Coordinate Framework.

(H) Location of all lenses from experimental animals registered to Allen Mouse Brain Common Coordinate Framework. Blue line indicates location of lens in (A). The dotted black line represents approximate location of tissue that was too damaged to reconstruct an accurate lens track. The white dotted line indicates PL borders.

(I) ML and DV coordinates of all neurons recorded in one example session, colored by region, and spike raster from example PL neurons.

(J) ROI masks for identified neurons and fluorescence traces from 5 example neurons.

Using Neuropixels 1.0 and 2.0 probes (Jun et al., 2017; Steinmetz et al., 2021), we recorded the activity of individual neurons in PFC, including anterior cingulate area (ACA), frontal pole (FRP), prelimbic area (PL), infralimbic area (ILA), and orbital area (ORB) (Laubach et al., 2018; Wang et al., 2020b). We also recorded from: secondary motor cortex (MOs), including anterolateral motor cotex (ALM), which has a well-characterized role in licking (Chen et al., 2017); olfactory cortex (OLF), including dorsal peduncular area (DP), dorsal taenia tecta (TTd), and anterior olfactory nucleus (AON), which receive input from the olfactory bulb (Igarashi et al., 2012; Mori and Sakano, 2021); and striatum, including caudoputamen (CP) and nucleus accumbens (ACB) (Fig. 1E-F), which are major outputs of PFC (Heilbronner et al., 2016). In a separate group of mice, we performed longitudinal 2-photon imaging through a GRIN lens to track the activity of individual neurons in PL across several days of behavioral training (Fig. 1G-H). Both techniques permitted robust measurement of the activity of neurons of interest and generated complementary results (Figs. 1I-J, S2).

Graded cue and lick coding across the recorded regions

In the electrophysiology experiment, we isolated the spiking activity of 5332 individual neurons in regions of interest across 5 mice (449-1550 neurons per mouse, Figs. 2A, S3A). The activity of neurons in all regions exhibited varying degrees of modulation in response to the six trial types (Fig. 2B). Broadly, there was strong modulation on CS+ and CS50 trials that appeared to be common to both odor sets (Fig. S3B). Across regions, there was heterogeneity in both the magnitude and the timing of the neural modulation relative to odor onset (Fig. S3C).

Graded cue and lick coding across the recorded regions.

(A) Location of each recorded neuron relative to bregma, projected onto 1 hemisphere. Each neuron is colored by CCF region. Numbers indicate total neurons passing quality control from each region.

(B) Mean normalized activity of all neurons from each region, aligned to odor onset, grouped by whether peak cue activity (0 - 2.5 s) was above (top) or below (bottom) baseline in held out trials. Number of neurons noted for each plot.

(C) Example kernel regression prediction of an individual neuron’s normalized activity on an example trial.

(D) CS+ trial activity from an example neuron and predictions with full model and with cues, licks, and reward removed. Numbers in parentheses are model performance (fraction of variance explained).

(E) Coordinates relative to bregma of every neuron encoding only cues or only licks, projected onto one hemisphere.

(F) Fraction of neurons in each region and region group classified as coding cues, licks, reward, or all combinations of the three.

(G) Additional cue (left) or lick (right) neurons in region on Y-axis compared to region on X-axis as a fraction of all neurons, for regions with non-overlapping 95% confidence intervals (see Methods).

To quantify the relative contribution of cues and conditioned responding (licking) to the activity of neurons in each region, we implemented reduced rank kernel regression (Steinmetz et al., 2019), using cues, licks, and rewards to predict neurons’ activity on held out trials (Figs. 2C, S4A). To determine the contribution of cues, licks, and rewards to each neuron’s activity, we calculated unique variance explained by individually removing each predictor from the model and calculating the reduction in model performance (Fig. 2D).

We identified individual neurons encoding cues, licks, or rewards as those for which that predictor uniquely contributed to 2% or more of their variance (a cutoff with an estimated false positive rate of 0.02%, see Methods). Neurons encoding cues (24% of all neurons), licks (11%), or both (16%) were most common. Neurons with any response to reward (independent of licking) were rare (5%) (Horst and Laubach, 2013). Cue neurons were characterized by sharp responses aligned to odor onset; in contrast, lick neurons’ responses were delayed and peaked around reward delivery (Fig. S4B-C), consistent with the timing of licks (Fig. 1C). The activity of cue neurons on rewarded and unrewarded CS50 trials validated our successful isolation of neurons with cue but not lick responses (Fig. S4D). The spatial distributions of cue and lick cells were noticeably different (Fig. 2E). The differences could be described as graded across regions, with the most lick neurons in ALM, and the most cue neurons in olfactory cortex and ORB, though each type of neuron was observed in every region (Fig. 2F-G, S5). Thus, our quantification of task encoding revealed varying prioritization of cue and lick signaling across all regions.

Cue value coding is present in all regions

To expand upon our analysis identifying cue-responsive neurons, we next assessed the presence of cue value coding in this population. The three cue types (CS+, CS50, or CS−) in our behavioral task varied in relative value according to the predicted probability of reward (Eshel et al., 2016; Fiorillo et al., 2003; Winkelmeier et al., 2022). We reasoned that a neuron encoding cue value should have activity that scaled with the relative value of the cues (Fig. 3A). We modeled this relationship on a per-neuron basis by scaling a single cue kernel by the typical number of anticipatory licks for each cue (see Methods, Fig. 3B). This model describes cue activity as similar across odors of the same value, and scaling in magnitude according to each odor’s value. As a shuffle control, we also fit each neuron with 89 additional models containing all possible permutations of reassigning the original values across the six odors, as well as an ‘untuned’ model with the same value for all odors (Fig. S6). After removing neurons best fit by the untuned model, the remaining 90 models would be equally likely to fit each neuron best if cue responses were independent of cue association, as would be expected with a sensory olfactory code (Pashkovski et al., 2020; Stettler and Axel, 2009). We found, however, that the original model with cue responses that scale directly with value (CS+ > CS50 > CS−) was the best model for a large fraction of cue neurons (30%), far exceeding chance (1%) (Fig. 3C). We refer to these neurons as value cells (Fig. S7A). Interestingly, five additional models stood out as the best model for sizable fractions of cue neurons. These models corresponded to the five alternative rankings of cue types (for example, CS50 > CS− > CS+ rather than CS+ > CS50 > CS−) irrespective of odor set (Fig. 3D) and accounted for 22% of cue neurons. This population of cells, which we refer to as trial type cells (Fig. S7B), encoded each cue’s reward probability independent of its particular odor, but with firing rates not proportional to value. The difference between value and trial type cells, therefore, is that value cells have both CS− activity closer to baseline and CS+ and CS50 activity occurring along the same dimension relative to CS− activity, evident in a smaller angle between CS+ and CS50 population response vectors for value cells (Fig. S7C-E).

Widespread cue value coding.

(A) Normalized activity of an example value cell with increasing modulation for cues of higher value.

(B) For the same neuron, model-fit cue kernel for the original value model and with one of the 89 alternatively-permuted value models.

(C) Distribution of best model fits across all cue neurons. Light blue is value model, purple is trial type models, gray is untuned model, and the remaining models are dark blue.

(D) First principal component of all neurons best fit by the original value model or other trial type and untuned models.

(E) Fraction of neurons in each region and region group classified as value cells (blue) and other cue neurons (gray), as well as fraction (+/− 95% CI) estimated from a linear mixed effects model with random effect of session (see Methods). n.s. indicates overlapping 95% CI for all three region groups.

(F) Additional value cells in region on Y-axis compared to region on X-axis as a fraction of all neurons, for regions with non-overlapping 95% confidence intervals. * indicates non-overlapping 95% CI for all three region groups.

(G) Principal component most related to value for value cells from all regions (2nd component in ACA, 1st component in all others).

(H) Same as (F), for region groups.

(I) PL population activity 0 to 2.5 s from odor onset projected onto first 3 principal components, defined on odor set 1 activity.

(J) Normalized distance (+/− SD) from baseline firing in odor set 1 PCA space (first 3 components) for odor set 2 trial types for value cells (top) and other cue cells (bottom) for 5000 selections of neurons (see Methods).

The frequency of value cells was similar across the recorded regions (Fig. 3E). Indeed, despite the regional variability in number of cue cells broadly (Fig. 2F-G), there were no regions that statistically differed in their proportions of value cells (Figs. 3E, S8). Though there were fewest cue neurons in motor cortex, they were more likely than cue neurons in other regions to encode value, followed by PFC (Fig. S9). The frequency of trial type cells was more variable, peaking in DP (Fig. 3F) and decreasing from olfactory to PFC to motor cortex (Fig. S8).

We next investigated the robustness of the value representation in each of our recorded regions. Principal component analysis on value cells from each region revealed similarly strong value-related dynamics across motor, prefrontal, and olfactory regions (Fig. 3G-H). Within the population space defined by the first 3 principal components of odor set 1 activity of value cells in each region, odor set 2 activity of this population showed strong value features; specifically, population activity deflected furthest from baseline for CS+ trials, less for CS50 trials, and least for CS− trials, with similar robustness in ALM, FRP, PL, ILA, ORB, and TTd (Fig. 3I-J). Taken together, these data illustrate that, in contrast to cue and lick coding broadly, and in contrast to trial type cue coding, value coding is similarly represented across the regions we sampled. In fact, this observation extended to the striatal regions we sampled as well, indicating that such value coding is widespread even beyond cortex (Fig. S10).

Because cue valuations can be influenced by preceding reward outcomes, We next considered whether the cue value signaling we detected was sensitive to the history of reinforcement (Nakahara et al., 2004; Ottenheimer et al., 2020; Winkelmeier et al., 2022). To estimate the subjects’ trial-by-trial cue valuation, we fit a linear model predicting the number of anticipatory licks on each trial from cue type, overall reward history, and cue type-specific reward history; we used the model prediction of licks per trial as our estimate of value. We found a strong influence of cue type-specific reward history and a more modest influence of overall reward history (Fig. 4A). These effects of reward history on lick rate are also visible when comparing the lick rate on trials grouped by the model-estimated value (Fig. 4B). Thus our behavioral model revealed that subjects more highly valued trials preceded by rewards.

A subset of value cells incorporate reward history.

(A) Coefficient weight (+/− standard error from model fit) for reward outcome on the previous 10 trials of any type (left) and on the previous 10 trials of the same cue type (right) for a linear model predicting the number of anticipatory licks on every trial of every session. Lick rates were normalized so that the maximum lick rate for each session was equal to 1. Colored lines are models fit to each individual mouse.

(B) Mean (+/− SEM) lick rate across mice (n = 5 mice) on trials binned according to value estimated from the lick model, incorporating recent reward history.

(C) Normalized activity of an example history value cell with increasing modulation for cues of higher value.

(D) For the same neuron, model-predicted activity with the original value model (left) and with trial-by-trial value estimates from the lick model (right).

(E) The activity of all cells in each category projected onto the coding dimension maximally separating CS− and CS+ for CS50 trials binned by value estimated from the lick model.

(F) The mean (+/− std across 5000 bootstrapped selections of neurons) activity (1 - 2.5 s from odor onset) along the coding dimension maximally separating CS− and CS+ for CS50 trials binned by value estimated from the lick model. * = p < 10−7 comparing highest and lowest value CS50 trials (other categories p > 0.23).

(G) Fraction of neurons in each region and region group classified as history value (light blue), non-history value (blue), and other cue neurons (gray), as well as estimated fraction (+/−95% CI) with random effect of session (see Methods). * indicates non-overlapping 95% CI for PFC and olfactory regions.

(H) Fraction of value neurons in each region group with history effect and estimated fraction (+/− 95% CI) with random effect of session. * indicates non-overlapping 95% CI for olfactory compared to PFC and motor regions.

We therefore investigated whether value cells showed similar trial-by-trial differences in their firing rates (Fig. 4C). To test this, we compared the fit of our original value model (Fig. 3B) with an alternative model in which the kernel scaled with the per-trial value estimates from our lick regression (Fig. 4D). Overall, 30% of value cells were better fit by the history-dependent value model than by the model without history dependence. To further evaluate the history component, we calculated these neurons’ activity on CS50 trials of varying value and projected it onto the population dimension maximizing the separation between CS+ and CS−. We hypothesized that high value CS50 trials would be closer to CS+ activity while low value CS50 trials would be closer to CS− activity. Indeed, history value cells (and lick cells) demonstrated graded activity along this dimension, in contrast to non-history value cells and trial type cells (Fig. 4E-F). Finally, we examined the spatial distribution of history value cells and found similar, low numbers across all regions, but with higher prevalence overall in PFC than in olfactory cortex (Fig. 4G). Also, as a fraction of all value cells, history cells were least common in olfactory cortex (Fig. 4H), providing evidence for less dynamic task-centric cue signaling there.

Cue coding emerges along with behavioral learning

To determine the timescales over which these coding schemes emerged and persisted, we performed longitudinal 2-photon imaging and tracked the activity of individual neurons across several days of behavioral training (Fig. 5A). We targeted our GRIN lenses to PL, a location with robust cue and lick coding (Fig. 2F) and where cue responses predominantly encode cue trial type and value (Fig. 3E-F). Mice (n = 8) developed anticipatory licking on day 1 that differentiated CS+ trials from CS50 (t(7) = 3.2, p = 0.015) and CS− (t(7) = 7.0, p = 0.0002) trials and CS50 trials from CS− (t(7) = 3.7, p = 0.008) trials (Fig. 5B-C). Visualizing the normalized activity across the imaging plane following CS+ presentation early and late on day 1 revealed a pronounced increase in modulation in this first session (Fig. 5D-E). Individual neurons (n = 705, 41-165 per mouse) also displayed a notable increase in modulation in response to the CS+ after task learning (Fig. 5F).

Acquisition of conditioned behavior and cue encoding in PFC.

(A) Training schedule for 5 of the mice in the imaging experiment. An additional 3 were trained only on odor set 1.

(B) Mean (+/− SEM) licking on early (first 60) and late (last 60) trials from day 1 of odor set 1 (n = 8 mice).

(C) Mean (+/− SEM) baseline-subtracted anticipatory licks for early and late trials from each day of odor set 1. Thin lines are individual mice (n = 8 mice).

(D) Standard deviation of fluorescence from example imaging plane.

(E) Normalized activity of each pixel following CS+ presentation in early and late day 1 trials.

(F) Normalized deconvolved spike rate of all individual neurons for early and late trials on day 1.

(G) Proportion of neurons classified as coding cues, licks, rewards, and all combinations for each third of day 1.

(H) Mean(+/− SEM) unique variance explained by cues, licks, and rewards for neurons from each mouse. Thin lines are individual mice. Unique variance was significantly different across session thirds for cues (F (2, 21) = 3.71, p = 0.04) but not licks (F (2, 21) = 0.37, p = 0.69) or reward (F (2, 21) = 0.65, p = 0.53, n = 8 mice, one-way ANOVA).

(I) Mean (+/− SEM) normalized deconvolved spike rate for cells coding cues, licks, both, or neither on early and late trials, sorted by whether peak cue activity (0 - 2.5 s) was above (top) or below (bottom) baseline for late trials.

To determine whether this increase in activity was best explained by a cue-evoked response, licking, or both, we again used kernel regression to fit and predict the activity of each neuron for early, middle, and late trials on day 1. The number of individual neurons encoding cues more than doubled from early to late day 1 trials (Fig. 5G). The unique variance cues increased across this first session, in contrast to licks and reward (Fig. 5H). This stark change in cue coding was also noticeable when plotting neurons encoding cues, licks, or both, as defined at the end of the session, on both early and late trials (Fig. 5I). These data indicated that PFC neural activity related to cues (but not licks) rapidly emerges during initial learning of the behavioral task.

Cue and lick coding is stable across days

We next assessed whether cue and lick coding were stable across days. By revisiting the same imaging plane on each day of training, we were able to identify neurons that were present on all three days of odor set 1 training (n = 371, 20-65 per mouse) (Fig. 6A-B). There was remarkable conservation of task responding across days, both on an individual neuron level (Fig. 6C) and across all imaged neurons (Fig. 6D). To quantify coding stability, we fit our kernel regression to the activity of each neuron on day 3 (Fig. 6E) and then used these models to predict activity on early, middle, and late trials on days 1-3 (Fig. 6F). Day 3 model predictions were most highly correlated with true activity on day 3, but they outperformed shuffle controls at all time points except early day 1, demonstrating preservation of a learned coding scheme. We then asked more specifically whether cells coding cues, licks, and both maintained their coding preferences across days. For each group of cells, we calculated their unique cue, lick, and reward variance at each time point. The preferred coding of each group, as defined on day 3, was preserved in earlier days (Fig. 6G). Thus, cue and lick coding are stable properties of PFC neurons across multiple days of behavioral training.

Cue and lick coding is stable across days.

(A) Standard deviation fluorescence from example imaging plane.

(B) Masks (randomly colored) for all tracked neurons from this imaging plane.

(C) Deconvolved spike rate on every CS+ trial from all three days of odor set 1 for an example neuron. Vertical dashed line is reward delivery. Color axis as in (D).

(D) Normalized deconvolved spike rate for all tracked neurons on all three days of odor set 1.

(E) Fraction of tracked neurons coding cues, licks, rewards, and their combinations on day 3.

(F) Model performance when using day 3 models to predict the activity of individual neurons across odor set 1 training, plotted as mean (+/− SEM) correlation between true and predicted activity across mice. Thin lines are individual mice. Performance was greater than shuffled data at all time points (p < 0.0001) except early day 1 (p = 0.21, Bonferroni-corrected, n = 8 mice).

(G) Mean (+/− SEM) unique cue, lick, and reward variance for cells classified as coding cues, licks, both, or neither on day 3. Day 3 cue cells had increased cue variance on day 2 (p < 10−7, see Methods) and 1 (p < 0.03) relative to lick and reward variance. Same pattern for lick cells on day 2 (p < 0.0001) and day 1 (p < 0.01).

A subset of mice (n = 5) also learned a second odor set, presented on separate days. Activity was very similar for both odor sets, evident across the entire imaging plane (Fig. 7A), for individual tracked neurons (n = 594, 81-153 per mouse) (Fig. S2B), and for kernel regression classification of these neurons (Fig. 7B). Notably, odor set 1 models performed similarly well at predicting both odor set 1 and odor set 2 activity (Fig. 7C). Moreover, cue, lick and both neurons maintained their unique variance preference across odor sets (Fig. 7D). Finally, to investigate the presence of value coding across odor sets over separate days, we fit tracked cue neurons with the value model and its shuffles. Even with odor sets imaged on separate days, we again found that the value model was the best model for a sizable fraction (27%) of cue neurons, demonstrating that value coding is conserved across stimulus sets and days (Fig. 7E).

Cue and lick coding in separately trained odor sets.

(A) Normalized activity of all pixels in the imaging plane following CS+ presentation on day 3 of each odor set.

(B) Fraction of neurons coding for cues, licks, rewards, and their combinations for day 3 of each odor set.

(C) Mean(+/− SEM, across mice) correlation between activity predicted by odor set 1 models and true data, for real (black) and trial shuffled (gray) activity. Thin lines are individual mice. F (1, 16) = 3.2, p = 0.09 for main effect of odor set, F (1, 16) = 135, p < 10−8 for main effect of shuffle, F (1, 16) = 2.2, p = 0.16 for interaction, n = 5 mice, two-way ANOVA.

(D) Mean(+/− SEM, across mice) unique cue, lick, and reward variance for cells classified as coding cues, licks, both, or neither for odor set 1. For each category, odor set 1 unique variance preference was maintained for odor set 2 (p < 0.04) except for both cells, for which lick and reward variance were not different in odor set 2 (p = 0.22, Bonferroni-corrected, n = 5 mice).

(E) Distribution of best model fits across all cue neurons, and the first principal component of the activity of all neurons best fit by the top 3 models.

Discussion

Our experiments assessed how coding for reward-predicting cues and reward-seeking actions differed across brain regions and across multiple days of training. We found coding for cues and licks in all regions we sampled, but their proportions varied in a graded way across those regions. In contrast to regional differences in the proportion of cue-responsive cells, cue value coding was similarly represented in all regions. Coding for cue value was robust, occurring with far greater frequency than chance, and, in a subset of neurons, incorporated the recent reward history. Cue coding was established within the first day of training and neurons encoding cues or licks maintained their coding preference across multiple days of the task. These results demonstrate a lack of regional specialization in value coding and the stability of cue and lick codes in PFC.

Graded cue and lick coding across regions

We found robust and separable coding for licks and cues (and combined coding of both) in all regions using electrophysiology and in PL using calcium imaging. The widespread presence of lick coding is consistent with recent reports of distributed movement and action coding (Musall et al., 2019; Steinmetz et al., 2019; Stringer et al., 2019); however, we saw sizable differences in the amount of lick coding across recorded regions. Notably, ALM had the greatest number of lick neurons, as well as the fewest cue neurons, perhaps reflecting its specialized role in the preparation and execution of licking behavior (Chen et al., 2017). Conversely, the olfactory cortical regions DP, TTd, and AON had the most cue neurons (especially non-value coding cue neurons), suggesting a role in early odor identification and processing (Mori and Sakano, 2021). PFC subregions balanced lick and cue coding, consistent with their proposed roles as association areas (Klein-Fluügge et al., 2022; Miller and Cohen, 2001), but there was variability within PFC as well. In particular, ORB had a greater fraction of cue cells than any other subregions, consistent with its known dense inputs from the olfactory system (Ekstrand et al., 2001; Price, 1985; Price et al., 1991). Thus, our results establish that the neural correlates of this Pavlovian conditioned behavior consist of a gradient of cue and response coding rather than segmentation of sensory and motor responses.

Widespread value signaling

Value signals can take on many forms and occur throughout task epochs. In our experiments, we focused on the predicted value associated with each conditioned stimulus, which is crucial for understanding how predictive stimuli produce motivated behavior (Berridge, 2004). Most comparisons of single neuron stimulus-value signaling across PFC have been conducted in primates. These studies have found neurons correlated with stimulus-predicted value in many subregions, with the strongest representations typically in ORB (Hunt et al., 2018; Kennerley et al., 2009; Roesch and Olson, 2004; Sallet et al., 2007). In rodents, there is also a rich history of studying value signaling in ORB (Kuwabara et al., 2020; Namboodiri et al., 2019; Schoenbaum et al., 2003; Stalnaker et al., 2014; Sul et al., 2010; van Duuren et al., 2009; Wang et al., 2020a), but there have been many reports of value-like signals in frontal cortical regions beyond ORB, as well (Allen et al., 2019; Kondo and Matsuzaki, 2021; Otis et al., 2017; Wang et al., 2020a). In our present experiment, we sought to expand upon the results from the rodent literature by separating cue activity from licking, which can track with value and may confound interpretation of the signal, by including more than two cue types, which provided a rich space to assess value coding, and by sampling from many frontal regions in the same experiment.

When considering the number of neurons responsive to cues rather than licks, our data confirmed the importance of ORB, which has more cue-responsive neurons than motor and other prefrontal regions. However, by analyzing the activity of cue-responsive neurons across all 6 odors predicting varying probabilities of reward, we were able to separate out neurons coding value from other neurons, which included a population that had consistent responses for odors with the same associated reward probability (trial type) but activity that did not scale according to probability, consistent with the possibility of nonlinear value coding (Enel et al., 2021). When only considering cue neurons with linear coding of value, the distribution was even across regions. One consequence of a widely distributed value signal is that manipulating only one subregion would be less likely to fully disrupt value representations, which is consistent with the results of studies comparing functional manipulations of different PFC subregions (Chudasama and Robbins, 2003; Dalton et al., 2016; St. Onge and Floresco, 2010; Verharen et al., 2020; Wang et al., 2020a). Different subregional impacts on behavior may reveal biases in how the value signal in each region contributes to reward-related behaviors, for instance during learning or expression of a reward association (Namboodiri et al., 2019; Otis et al., 2017; Wang et al., 2020a). A related interpretation is that, in this task, there may be other properties that correlate with cue value, and the homogeneous value representation we observed across regions masks regional differences in tuning to these other correlated features, such as motivation (Roesch and Olson, 2004) and a host of related concepts, including salience, uncertainty, vigor, and arousal (Hayden and Niv, 2021; Stalnaker et al., 2015; Zhou et al., 2021), which can have different contributions to behavior. This interpretation is consistent with broader views that observations of ‘value’ signals are often misconstrued (Zhou et al., 2021) and that pure abstract value may not be encoded in the brain at all (Hayden and Niv, 2021). Although the identification of value in our task was robust to three levels of reward probability across two stimulus sets, the fact that this signal was widespread contributes to the case for revisiting the definition and interpretation of value to better understand regional specialization.

In our analysis, we uncovered a distinction between neurons encoding the overall value of cues and those with value representations that incorporated the recent reward history. Neurons with history effects were rarer but also widespread. These neurons may have a more direct impact on behavioral output in this task, because the lick rate also incorporated recent reward history. Notably, the impact of reward history on these neurons was noticeable even prior to cue onset, consistent with a previously proposed mechanism for persistent value representations encoded in the baseline firing rates of PFC neurons (Bari et al., 2019).

Given the presence of value coding in olfactory cortex, the question remains of where odor information is first transformed into a value signal. In fact, there have been multiple reports of some association-related modification of odor representations as early as the olfactory bulb (Chu et al., 2016; Doucette et al., 2011; Koldaeva et al., 2019; Li et al., 2015). Considering the prevalence of value and non-value trial type coding we observed in AON, DP, and TTd, perhaps these regions are a crucial first step in processing and amplifying task-related input from the olfactory bulb. Because they provide input to PFC (Bhattarai et al., 2021; Igarashi et al., 2012), they may be an important source of the odor coding we observed there. Previous recordings in AON, DP, and TTd were in anesthetized rodents (Cousens, 2020; Kikuta et al., 2008; Lei et al., 2006; Tsuji et al., 2019); as the first recordings in awake behaving animals, our results bring these regions into focus for future work on the transformation of odor information into task-relevant coding.

Stability of PFC codes

Previous reports have observed drifting representations in PFC across time (Hyman et al., 2012; Malagon-Vina et al., 2018), and there is compelling evidence that odor representations in piriform drift over weeks when odors are experienced infrequently (Schoonover et al., 2021). On the other hand, it has been shown that coding for odor association is stable in ORB and PL, and that coding for odor identity is stable in piriform (Wang et al., 2020a), with similar findings for auditory Pavlovian cue encoding in PL (Grant et al., 2021; Otis et al., 2017) and ORB (Namboodiri et al., 2019). We were able to expand upon these data in PL by identifying both cue and lick coding and showing separable, stable coding of cues and licks across days and across sets of odors trained on separate days. We were also able to detect value coding common to two stimulus sets presented on separate days. This consistency in cue and lick representations indicates that PL serves as a reliable source of information about cue associations and licking during reward seeking tasks, perhaps contrasting with other representations in PFC (Hyman et al., 2012; Malagon-Vina et al., 2018). Interestingly, the presence of lick, but not cue coding at the very beginning of day 1 of training suggests that lick cells in PL are not specific to the task but that cue cells are specific to the learned cue-reward associations.

Overall, our work emphasizes the importance of evaluating regional specialization of neural encoding with systematic recordings in many regions using the same task. Future work will clarify whether cue value is similarly widely represented in other reward-seeking settings and whether there are regional differences in the function of the value signal.

eLife Assessment

This study makes valuable observations about the representation of "value" in the mouse brain by using a nice task design and recording from an impressive number of brain regions. The combination of state-of-the-art imaging and electrophysiology data offer solid support for the authors' conclusions. If more thorough statistical analysis of the response of neuronal populations supports the claims, the paper will be of interest to a broad audience of neuroscientists interested in reward processing in the brain.

Consensus Public Review:

Ottenheimer et al., present an interesting study looking at the neural representation of value in mice performing a pavlovian association task. The task is repeated in the same animals using two odor sets, allowing a distinction between odor identity coding and value coding. The authors use state-of-the-art electrophysiological techniques to record thousands of neurons from 11 frontal cortical regions to conclude that 1) licking is represented more strongly in dorsal frontal regions, 2) odor cues are represented more strongly in ventral frontal regions, 3) cue values are evenly distributed across regions. They separately perform a calcium imaging study to track coding across days and conclude that the representation of task features increments with learning and remains stable thereafter.

Overall, these conclusions are interesting and mostly well supported by the data, although there are some doubts about their definition of value coding. One limitation is the lack of focus on population-level dynamics from the perspective of decoding, with the analysis focusing primarily on encoding analyses within individual neurons.

Some specific comments:

The authors use reduced-rank kernel regression to characterize the 5332 recorded neurons on a cell-by-cell basis in terms of their responses to cues, licks, and reward, with a cell characterized as encoding one of these parameters if it accounts for at least 2% of the observed variance. At least 50% of cells met this inclusion criterion in each recorded area. 2% feels like a lenient cutoff, and it is unclear how sensitive the results are to this cutoff, though the authors argue that this cutoff should still only allow a false positive rate of 0.02% (determined by randomly shuffling the onset time of each trial).

Having identified lick, reward, and cue cells, the authors next select the 24% of "cue-only" neurons and look for cells that specifically encode cue value. Because the animal's perception of stimulus value can't be measured directly, the authors created a linear model that predicts the amount of anticipatory licking in the interval between odor cue and reward presentations. The session-average-predicted lick rate by this model is used as an estimate of cue value and is used in the regression analysis that identified value cells. (Hence, the authors' definition of value is dependent on the average amount of anticipatory behavior ahead of a reward, which indicates that compared to the CS+, mice licked around 70% as much to the CS50 and 10% as much to the CS-.) The claim that this is an encoding of value is strengthened by the fact that cells show similar scaling of responses to two odor sets tested. Whereas the authors found more "lick" cells in motor regions and more "cue" cells in sensory regions, they find a consistent percentage of "value" cells (that is, cells found to be cue-only in the initial round of analysis that is subsequently found to encode anticipatory lick rate) across all 11 recorded regions, leading to their claim of a distributed code of value.

In subsequent sections, the authors expand their model of anticipatory-licking-as-value by incorporating trial and stimulus history terms into the model, allowing them to predict the anticipatory lick rate on individual trials within a session. They also use 2-photon imaging in PFC to demonstrate that neural coding of cue and lick are stable across three days of imaging, supported by two lines of evidence. First, they show that the correlation between cell responses on all periods except for the start of day 1 is more correlated with day 3 responses than expected by chance (although the correlation is still quite low, for example, 0.2 on day 2). Second, they show that cue identity is able to capture the highest unique fraction of variance (around 8%) in day 3 cue cells across three days of imaging, and similarly for lick behavior in lick cells and cue+lick in cue+lick cells. Nonetheless, their sample rasters for all imaged cells also indicate that representations are not perfectly stable, and it will be interesting to see what *does* change across the three days of imaging.

Importantly, the authors do not present evidence that value itself is stably encoded across days, despite the paper's title. The more conservative in its claims in the Discussion seems more appropriate: "these results demonstrate a lack of regional specialization in value coding and the stability of cue and lick [(not value)] codes in PFC."

Author Response:

The following is the authors' response to the original reviews.

We thank the editor and reviewers for their careful consideration of our manuscript and very helpful feedback, which guided us in improving our manuscript. We would like to highlight three main areas of improvement in this version:

Statistical rigor: we have added more detail to justify our 2% cutoff for GLM variable coding, implemented stricter shuffling and cutoffs for value and history coding, and provided more information on the statistical significance of our pairwise comparisons across regions and groups. These go well beyond the field standard for identifying and comparing neural encoding of task features.

Identification of value coding: we have implemented reviewer suggestions about kernel regression and value coding shuffles, providing even stronger evidence that value signaling among cue neurons is more prevalent than expected by chance, more prevalent than any other cue coding patterns, and present in all recorded regions. The rigor of this analysis is only possible due to our unique task design with 6 cues across two stimulus sets, and our consideration of 153 possible coding models exceeds standard practice for identifying value signals. We now implement population decoding, as well, providing additional support for a robust and widely-distributed value code.

Stability of value code: we have updated our terminology to better highlight that the value signals in our imaging dataset are indeed identified across days, and we add new analysis to show conservation of value-like signals across training days.

Thanks to the reviewers’ suggestions, our manuscript now has substantially stronger support for the presence of stable and distributed cue value signaling. We address the specific points below.

Excerpts from the Consensus Public Reviews:

One limitation is the lack of focus on population-level dynamics from the perspective of decoding, with the analysis focusing primarily on encoding analyses within individual neurons.

To address this limitation, we now include population-level decoding analysis (new panels, Figs. 3G-H, 4E). This new analysis reveals that, although value neurons can be used to decode cue identity on par with other cue cells, value neurons are more accurate at predicting the value of held out cues (never seen by the model), highlighting the utility of a value signal as a way to consistently represent the value of different stimulus sets.

Moreover, we find comparable value prediction performance when using value neurons from each region (Fig. 4E), adding more support for the similarity of this signal across regions:

The authors use reduced-rank kernel regression to characterize the 5332 recorded neurons on a cell-by-cell basis in terms of their responses to cues, licks, and reward, with a cell characterized as encoding one of these parameters if it accounts for at least 2% of the observed variance. At least 50% of cells met this inclusion criterion in each recorded area. 2% feels like a lenient cutoff, and it is unclear how sensitive the results are to this cutoff, though the authors argue that this cutoff should still only allow a false positive rate of 0.02% (determined by randomly shuffling the onset time of each trial.)

We have provided more information about the 2% cutoff in a new figure, Figure 2-figure supplement 3. We reanalyzed the false positive rate and found that at a cutoff of 2% (but not 0.5% or 1%) there were no false positives (Figure 2-figure supplement 3B). Thus, we are confident that all neurons contain true task-related signals. Moreover, we found that the pattern of results remains largely unchanged as we change the cutoff over a range from 0.5% to 5%. With more stringent cutoffs, we begin to lose neurons with robust task-related responses (Figure 2-figure supplement 3E), so we continue to use the 2% cutoff in this version of the manuscript.

First, they show that the correlation between cell responses on all periods except for the start of day 1 is more correlated with day 3 responses than expected by chance (although the correlation is still quite low, for example, 0.2 on day 2).

We agree that a correlation of 0.2 does not seem like a large effect, however the variability in neuronal responses and noise level of the measurement enforce a ceiling that we can estimate by predicting data from the same session that it was trained on. We have replotted these data (new panel Fig. 7G) with the correlation normalized to the cross-validated performance on the training day’s data. This shows that the models do about half as well in session 1 and session 2 compared to session 3. The original plot is in a new supplementary figure, Figure 7-figure supplement 1B.

To further emphasize the similarity across days, we have added new panels (Fig. 7E and Figure 7-figure supplement 1A) showing that, across mice, a typical neuron was more correlated with its own activity on the subsequent day than with ~90% of the other neurons (shuffle controls, 50%).

Second, they show that cue identity is able to capture the highest unique fraction of variance (around 8%) in day 3 cue cells across three days of imaging, and similarly for lick behavior in lick cells and cue+lick in cue+lick cells. Nonetheless, their sample rasters for all imaged cells also indicate that representations are not perfectly stable, and it will be interesting to see what *does* change across the three days of imaging.

We agree that the representations are not perfectly stable and that is an interesting point of further investigation. One difference we did observe is increased cue coding across training (Figs. 6H, 7H).

Importantly, the authors do not present evidence that value itself is stably encoded across days, despite the paper's title. The more conservative in its claims in the Discussion seems more appropriate: "these results demonstrate a lack of regional specialization in value coding and the stability of cue and lick [(not value)] codes in PFC."

Due to confusing terminology on our part, the reviewers were mistaken about the timing of the experiment where we assess the stability of value coding. In the imaging sessions, odor sets were always presented on separate days. Thus, when we identify value coding in our imaged population, it is across two consecutive days with different odor sets, which is in itself evidence of a stable value code. We have updated our terminology and the text to make this clearer. We also added a new set of plots (Fig. 8H-I) showing the conservation of value-like signaling in cells we tracked across the first three sessions of odor set A, and, as above, that the correlation of these neurons across days is greater than expected by chance. These analyses lend further support to the stability of the value signal.

Additional technical comments:

1. The "shuffle #33" in figure 3B is confusing. The fit kernel in this shuffle shows that the "high" and "medium" responses increase above the pre-stimulus baseline. The "high" response is a combination of set 2 CS+ and set 1 CS50, both of which strongly suppressed the cell's firing over the 2.5-second window shown. Why then does the cue kernel fit these two trials predict an increase in firing rate above baseline at the 2.5-second time point? Is it a consequence of the reduced rank regression process, and if so, how? This strange-looking fit that does not well capture the response of the original cell makes me worry that the high fraction of identified "value" cells may be due to some constraint on the shuffle fits that leads them to often perform poorly.

To address this concern, we refit the value shuffle and its models using a full kernel regression model (rather than reduced ranks). It does improve the appearance of the kernel fits (updated Fig. 3B), and we now use this new approach when fitting cue coding models in the revised manuscript. The regularization inherent in reduced rank constrains the shape of the cue kernel somewhat, which contributed to the shape of the fits (although this did not negatively impact the variance explained); however, because of the importance of the shape of these alternative cue coding models to the interpretation of the analysis, we agree with the reviewers that this was worth improving. The main constraint on the value model and its shuffles, however, is that all cues must use the same template, scaled according to particular values assigned to each cue in each shuffle, which will doubtless lead to compromised (and strange-looking) fits when the shuffled values do not match the ranking of neuron’s cue activity. Critically, this constraint is applied equally to the value model and all the shuffles and would not bias the fits of any one model.

1. The "shuffle" condition when testing for value cells always assumes two high responses, two medium responses, and two low responses. This strategy doesn't account for cells that respond to only a subset of cues, as one might expect in a sparse-coding olfactory region. We suggest adding a set of shuffles where responses are split into two groups, with either 3 conditions per group or 2 in one group and 4 in the other.

We appreciate this valuable suggestion. We added all permutations of models with high responses to 6, 5, 4, 3, 2, or 1 odor cue to the analysis. We still find that the value model is the most frequent best model, displayed in new panels Fig. 3C-D and Figure 3-figure supplement 1A-B. The additional models allowed us to identify other neurons with cue activity best fit by models highly correlated with the ranked value model, which we term “value-like” neurons, including most neurons previously described as “trial-type” neurons. All 153 models and the fraction of neurons best fit by each one are depicted in Figure 3-figure supplement 1.

After implementing the changes to both the method of model fitting (full kernel regression, as noted above) and the possible alternative models, the distribution of value cells has changed slightly. All regions contain value cells, supporting our original conclusion that the value signal is distributed, but there is slight enrichment in PFC when combining these five regions together (Fig. 4A).

We have updated the conclusions of the paper accordingly:

Introduction: “Unexpectedly, in contrast to the graded cue and lick coding across these regions, the proportion of neurons encoding cue value was more consistent across regions, with a slight enrichment in PFC but with similar value decoding performance across all regions.”

Results: “Interestingly, the frequency of value cells was similar across the recorded regions (Fig. 4A). Indeed, despite the regional variability in number of cue cells broadly (Fig. 2F-G), there were very few regions that statistically differed in their proportions of value cells (Fig. 4A, Figure 4-figure supplement 1). Overall, though, there were slightly more value cells across all of PFC than in motor and olfactory cortex (Figs. 4A, Figure 4-figure supplement 1). Although there were the most cue neurons in olfactory cortex, these were less likely to encode value than cue neurons in other regions (Figure 4-figure supplement 2). Value-like cells were also widespread; they were less frequent in motor cortex as a fraction of all neurons, but they were equivalently distributed in all regions as a fraction of cue neurons (Fig. 4B, Figure 4-figure supplement 1, Figure 4-figure supplement 2).”

Discussion: “In contrast to regional differences in the proportion of cue-responsive neurons, cue value cells were present in all regions and could be used to decode value with similar accuracy regardless of region.” AND “The distribution of cue cells with linear coding of value was mostly even across regions, with slight enrichment overall in PFC compared to motor and olfactory cortex, but no subregional differences in PFC. Importantly, cue value could be decoded from the value cells in all regions with similar accuracy.”

1. On pages 11-12, the authors write "value coding is similarly represented across the regions we sampled." I feel this isn't quite what was shown: the authors have shown that all recorded regions contain a roughly comparable number of individual cells that are modulated by value, i.e. "value cells". However, the authors also showed that some recorded cells have mixed selectivity for value and other factors- it is possible that these mixed selectivity cells do vary between brain regions in their quantity or degree of value coding. Regions could potentially also vary in the dynamics of their value response, or in the trial-to-trial variability in the activity of value cells. I suggest the authors revise their original statement, for example by writing "we find a similar proportion of value-specific cells across the regions we sampled."

We thank the reviewer for carefully reviewing our claims. In addition to showing similar proportions of value cells, we also show that the value-related activity is similar (by plotting the first principal component of value and value-like cells, Fig. 4C-D) and that cue value could be decoded from the value cells in all regions with similar accuracy (new panel, Fig. 4E). We have updated the text to more accurately reflect these observations:

“In contrast to regional differences in the proportion of cue-responsive neurons, cue value cells were present in all regions and value could be decoded from them with similar accuracy regardless of region.”

1. We appreciate the authors' idea to introduce a history term to their value cell model but worry that the distinction between history-dependent value cells and lick/cue+lick cells in Figure 4 has gotten fuzzy. At this point, history-dependent value cells are the product of a set of steps: 1) they are identified as "cue" neurons because the cue type accounts for at least 2% of the variance, while the lick rate does not, then 2) among the cue neurons, a subset are identified as "value" neurons because their activity scales with the cue type across both odor sets, and then 3) among value neurons, the "history-dependent" value neurons show a response rate that scales with a model that predicts anticipatory licking. Our concern comes down to this: your conclusion that these cells are not licking cells hinges on the initial point that licking does not account for 2% of the observed variance in cell activity. But if you had dedicated an equal number of model parameters and selection steps to your licking model, might it still not turn out that a licking model predicts their activity as well as the history-dependent cue value model?

What would bolster our confidence here would be a comparison of variance explained: if you compare the predictions of the history-dependent value-encoding cue neuron model to the predictions of a simple lick neuron model, how much better does the former predict what the cells are doing? Are all those extra parameters and selection steps really contributing to an improved description of how neurons will respond?

First, we would like to emphasize that “cue” neurons, as a population, have no discernible modulation by licks, which can be seen when comparing their activity on CS50 trials with and without reward, when licking clearly varies (Figure 2-figure supplement 2D). A new panel, Figure 5E now depicts the improvement in variance explained by the history model over a lick only model. The improvement is robust and universal. This is because even though the number of anticipatory licks per trial is used to fit the weights of our trial value model, these cue neurons have temporal dynamics that are more consistent with cue presentation than the presence of licks. We explain more below in our response to point 7.

1. The paper's title claims that the coding of cue value is both stable and distributed. While the point for value coding being distributed is well supported with analysis, the claim that cue value coding is "stable" is weaker. The authors show in Figure 6 that cue identity best accounts for unique variance among cue cells across three days of imaging, but it does not follow that cue value is similarly stable. Figure 7 shows that on day 3 of imaging, the two odor sets have similar encoding- but this analysis is only performed within day 3, not across days. Why not examine unique variance among value cells over days, as was done for a cue, lick, and both cells in Figure 6G? That seems to be an important missing piece and a logical next step. The Discussion is more conservative in its claims- "these results demonstrate a lack of regional specialization in value coding and the stability of cue and lick [(not value)] codes in PFC." But this subtlety is missing from the paper's title and introduction.

First, an important correction. “This analysis is only performed within day 3, not across days,” is a misunderstanding of our experiment brought on by our confusing terminology, which we have updated. This figure (now Figure 8) analyzes two sessions performed on consecutive days: Odor Set A day 3 (A3) and Odor Set B day 3 (B3), which constitute days 5 and 6 of our experiment (see updated panels Fig. 1B, 6A). This is why identifying value signaling across both of these sessions is justification for a stable code; by definition, it was present on two consecutive days.

A limitation of our imaging experiment prevents us from evaluating value signaling in each individual session (like we did for cues and licks). For the imaging, we only presented one odor set per session (unlike the electrophysiology, where odor sets were presented in blocks). Our method of identifying value signals relies on two odor sets, so we cannot quantify it on a per session basis in the imaging. However, to address this as best we could, we identified CS+-preferring cue cells in session A3 (odor set A day 3) and plotted them for sessions A1-A3 (Fig. 8H), which reveals a conserved value-like signal across days. We also found that the correlation of the activity of these neurons across days was higher than expected by chance (Fig. 8I).

We have edited the discussion text about coding stability, adding in more detail and caveats:

“Previous reports have observed drifting representations in PFC across time (Hyman et al., 2012; Malagon-Vina et al., 2018), and there is compelling evidence that odor representations in piriform drift over weeks when odors are experienced infrequently (Schoonover et al., 2021). On the other hand, it has been shown that coding for odor association is stable in ORB and PL, and that coding for odor identity is stable in piriform (Wang et al., 2020a), with similar findings for auditory Pavlovian cue encoding in PL (Grant et al., 2021; Otis et al., 2017) and ORB (Namboodiri et al., 2019). We were able to expand upon these data in PL by identifying both cue and lick coding and showing separable, stable coding of cues and licks across days and across sets of odors trained on separate days. We were also able to detect value coding common to two stimulus sets presented on separate days, and conserved value features across the three training sessions. Notably, the model with responses only to CS+ cues best fit a larger fraction of imaged PL neurons than the ranked value model, a departure from the electrophysiology results. It would be interesting to know if this is due to a bias introduced by the imaging approach, the slightly reduced CS50 licking relative to CS+ licking in the imaging cohort, or the shorter imaging experimental timeline.

The consistency in cue and lick representations we observed indicates that PL serves as a reliable source of information about cue associations and licking during reward seeking tasks, perhaps contrasting with other representations in PFC (Hyman et al., 2012; Malagon-Vina et al., 2018). Interestingly, the presence of lick, but not cue coding at the very beginning of the first session of training suggests that lick cells in PL are not specific to the task but that cue cells are specific to the learned cue-reward associations. Future work could expand upon these findings by examining stimulus-independent value coding within session across many consecutive days.”

1. Considering licking as the readout of value has pros and cons. Anticipatory licking may be correlated with subjective value, but certainly nonlinearly. After all, licking has a ceiling and floor (bounded rate from 0->10 Hz). Are results consistent with the objective value of the cues (which are 0, .5, 1)? Which measure better explained the data?

Thanks to this important suggestion, we tried fitting another set of models with 0, 0.5, 1 as the cue values. We found the same pattern of results. Overall, the fits were slightly better with 0, 0.5, 1, with 50.6% of potential value neurons (found with either version of the model) better fit by 0, 0.5, 1, and with mean variance explained of 0.265 with 0, 0.5, 1 (compared to 0.264 with the anticipatory lick values). Without strong evidence to choose one model over the other, we decided to use 0, 0.5, 1 because it exactly reflects reward probability, and is more objective as the reviewer notes, whereas before we relied on a noisier estimate of subjective value. We have changed the text accordingly.

1. How can a neuron encode "Cue" in a value-dependent manner and not also encode licking, given they are correlated? If the kernel window includes anticipatory licking, and anticipatory licking is by definition related to value, then how could a licking kernel not at least explain some of that neuron's variance?

The trial estimates of value from the lick linear regression are derived from typical licking patterns across all sessions and do not incorporate the particular number of licks on a given trial or the latency of licking relative to cue onset. Although the trial value model is predicting the number of licks on each trial, it only uses cue identity and reward history to make its prediction, so it is not tightly correlated with the stochastic licks on a given trial. And, importantly, we input the trial value as a cue kernel spanning the entire cue period, whereas lick kernels, per our definition, are restricted to a window around when licking occurs, which generously encompasses neural signals relating to both lick initiation and feedback. Licking can explain some of value and (history) neurons’ variance, which you can see in our new panel Fig. 5E, but it does not contribute any unique variance to the model. That is, with or without licks, the model performs just as well, so the activity of the neuron does not track any of the unique features of licks over cues (like whether or not the mouse licked on trial, when the mouse started licking on a given trial). Without cues, however, the model does worse, which means that the neuron’s activity is modulated by cues separately from when the mouse is licking. Thus, we can conclude the neuron encodes cues, but we have no evidence the neuron encodes licks (beyond the extent to which licks are correlated with cues). In our example fit in 5E, you can see how, although licks track value, they cannot recapitulate the temporal dynamics of this cue neuron. We added more description of this distinction in the manuscript.

1. The ordering analysis with the 89 permutations is very nice for showing across the population the "value ordered" gains are the best explanation of the neural activity. However, it doesn't tell you that any one neuron significantly encodes value, or the strength of this effect if they do. For the former, they could compare to a null distribution of shuffled order of neural vs CS data, and consider neurons for which model is better than chance ( a .05 FDR on a null distribution would be appropriate). This is important for supporting their conclusion of the fraction of neurons encoding value for each region.

In fact, with so many alternative models, the probability of a neuron being best fit by the value model but not encoding value above chance is extremely low. To confirm this, we ran the reviewer’s suggested shuffle analysis, and found that 100% of value neurons performed above the 0.05 FDR. We have added this result to the methods:

“To verify the robustness of value coding in the neurons best fit by the ranked value model, we fit each of those neurons with 1000 iterations of the cue value model with shuffled cue order to create a null distribution. The fits of the original value model exceeded the 98th percentile of the null for all value neurons.”

1. Similarly the 65% cutoff for trial history relative to shuffled is unusually low and therefore not convincing these neurons significantly encode the value. Usually, 95% or 99% is selected to give you a more standard significance criterion (FDR).

We have changed the cutoff to 95%. We originally selected 65% because neurons in the 65% to 95% range had clear history effects, especially at the population level, but we appreciate the importance of rigorous selection. Note this shuffle is very strict, preserving CS+, CS50, CS- ranking but shuffling within-cue fluctuations in value due to trial history. With the stricter value and history shuffling, we now observe fewer history neurons, and they are most prevalent in PFC (Fig. 5I)

1. "Regions with non-overlapping CIs were considered to have significantly different fractions of neurons of that coding type." This isn't a statistical test. Confidence intervals are not the same as significance.

We now perform Bonferroni-corrected pairwise contrasts between all regions in the generalized linear mixed effects model. We added the p-values for all the comparisons that previously relied on non-overlapping confidence intervals in supplementary tables.

Minor comments:

The methods are hard to read. Most of the information seems to be there but in general, paragraphs need to be read over multiple times for meaning to emerge.

We have edited for clarity, and if there are particular sections that remain unclear, we would be happy to know which ones.

Why is there a block predictor in the encoding model?

Because not every odor is present in every block, we did not want our models to use the specific cue predictors to try to account for differences in baseline activity that naturally occur across the session. Thus, each of the six blocks has its own predictor that serves as a constant that can adjust for changing baseline firing rate. Importantly, the block predictor simply marks the passage of blocks and contains no information about the odors present. We added more information about this to the methods:

“For electrophysiology experiments, the model also included 6 constants that identified the block number, accounting for tonic changes in firing rate across blocks. Because not all cues were present in every block, this strategy prevented the cue kernels from being used to explain baseline changes across blocks.”

Did you use an elastic net rather than a lasso? What is the alpha parameter for lasso?

We used an elastic net with alpha = 0.5. We added this information to the methods.

Figure 3F legend doesn't seem to match the figure.

Corrected.

Version 2

Abstract

Summary

The ability to associate reward-predicting stimuli with adaptive behavior is frequently attributed to the prefrontal cortex, but the stimulus-specificity, spatial distribution, and stability of pre-frontal cue-reward associations are unresolved. We trained headfixed mice on an olfactory Pavlovian conditioning task and measured the coding properties of individual neurons across space (prefrontal, olfactory, and motor cortices) and time (multiple days). Neurons encoding cues or licks were most common in olfactory and motor cortex, respectively. By quantifying the responses of cue-encoding neurons to six cues with varying probabilities of reward, we unexpectedly found value coding in all regions we sampled, with some enrichment in prefrontal cortex. We further found that prefrontal cue and lick codes were preserved across days. Our results demonstrate that individual prefrontal neurons stably encode components of cue-reward learning within a larger spatial gradient of coding properties.

Introduction

Association of environmental stimuli with rewards and the subsequent orchestration of value-guided reward-seeking behavior are crucial functions of the nervous system linked to the pre-frontal cortex (PFC) (Klein-Flügge et al., 2022; Miller and Cohen, 2001). PFC is heterogeneous, with many studies noting subregional differences in both neural coding of (Hunt et al., 2018; Kennerley et al., 2009; Sul et al., 2010; Wang et al., 2020a) and functional impact on (Buckley et al., 2009; Dalley et al., 2004; Kesner and Churchwell, 2011; Rudebeck et al., 2008) value-based reward seeking in primates and rodents. Further, functional manipulations of PFC subregions exhibiting robust value signals do not always cause a discernible impact on reward-guided behavior (Chudasama and Robbins, 2003; Dalton et al., 2016; St. Onge and Floresco, 2010; Verharen et al., 2020; Wang et al., 2020a), encouraging investigation of differences between value signals across PFC. Within individual PFC subregions, multiple studies have observed evolving neural representations across time, calling into question the stability PFC signaling (Hyman et al., 2012; Malagon-Vina et al., 2018). A systematic comparison of coding properties across rodent PFC and related motor and sensory regions, as well as across days and stimulus sets, is necessary to provide a full context for the contributions of PFC subregions to reward processing.

Identifying neural signals for value requires a number of considerations. One issue is that other task features can vary either meaningfully or spuriously with value. In particular, action coding is difficult to parse from value signaling, given the high correlations between behavior and task events (Musall et al., 2019; Zagha et al., 2022) and widespread neural coding of reward-seeking actions (Steinmetz et al., 2019). Additionally, without a sufficiently rich value axis, it is possible to misidentify neurons as ‘value’ coding even though they do not generalize to valuations in other contexts (Hayden and Niv, 2021; Stalnaker et al., 2015; Zhou et al., 2021). Because reports of value have come from different experiments across different species, it is difficult to compare the presence of value signaling even across regions within prefrontal cortex (Hayden and Niv, 2021; Hunt et al., 2018; Kennerley et al., 2009; Namboodiri et al., 2019; Otis et al., 2017; Stalnaker et al., 2015; Sul et al., 2010; Wang et al., 2020a; Zhou et al., 2021).

In this work, we sought to address the existing ambiguity in the distribution and stability of value signaling. We implemented an olfactory Pavlovian conditioning task that permitted identification of value correlates within the domain of reward probability across two separate stimulus sets. With acute in vivo electrophysiology recordings, we were able to assess coding of this task across 11 brain regions, including five PFC subregions, as well as olfactory and motor cortex, in a single group of mice, permitting a well-controlled comparison of coding patterns across a large group of task-relevant regions in the same subjects. Unexpectedly, in contrast to the graded cue and lick coding across these regions, the proportion of neurons encoding cue value was more consistent across regions, with a slight enrichment in PFC but with similar value decoding performance across all regions. To assess coding stability, we performed 2-photon calcium imaging of neurons in PFC for multiple days and determined that the cue and lick codes we identified were stable over time. Our data demonstrate universality and stability of cue-reward coding in mouse cortex.

Results

Distributed neural activity during an olfactory Pavlovian conditioning task

We trained mice on an olfactory Pavlovian conditioning task with three cue (conditioned stimulus) types that predicted reward on 100% (‘CS+’), 50% (‘CS50’), or 0% (‘CS-’) of trials (Fig. 1A). Each mouse learned two odor sets (odor sets A and B), trained and imaged on separate days and then, for electrophysiology experiments, presented in six alternating blocks of 51 trials during recording sessions (Fig. 1B). Mice developed anticipatory licking (Fig. 1C-D), and the rate of this licking correlated with reward probability (Fig. S1), indicating that subjects successfully learned the meaning of all six odors.

Electrophysiology and calcium imaging during olfactory Pavlovian conditioning.

(A) Trial structure in Pavlovian conditioning task.

(B) Timeline for mouse training.

(C) Mean (+/− standard error of the mean (SEM)) lick rate across mice (n = 5) on each trial type for each odor set during electrophysiology sessions. CS50(r) and CS50(u) are rewarded and unrewarded trials, respectively. Inset: mean anticipatory licks (change from baseline) for the CS+ and CS50 cues for every session, color-coded by mouse. F (1, 66) = 36.6 for a main effect of cue in a two-way ANOVA including an effect of subject.

(D) Same as (C), for the third session of each odor set (n = 5 mice). t(4) = 5.4 for a t-test comparing anticipatory licks on CS+ and CS50 trials.

(E) Neuropixels probe tracks labeled with fluorescent dye (red) in cleared brain (autofluorescence, green). AP, anterior/posterior; ML, medial/lateral; DV, dorsal/ventral. Allen CCF regions delineated in gray. Outline of prelimbic area in purple.

(F) Reconstructed recording sites from all tracked probe insertions (n = 44 insertions, n = 5 mice), colored by mouse.

(G) Sample histology image of lens placement. Visualization includes DAPI (blue) and GCaMP (green) signal with lines indicating cortical regions from Allen Mouse Brain Common Coordinate Framework.

(H) Location of all lenses from experimental animals registered to Allen Mouse Brain Common Coordinate Framework. Blue line indicates location of lens in (A). The dotted black line represents approximate location of tissue that was too damaged to reconstruct an accurate lens track. The white dotted line indicates PL borders.

(I) ML and DV coordinates of all neurons recorded in one example session, colored by region, and spike raster from example PL neurons.

(J) ROI masks for identified neurons and fluorescence traces from 5 example neurons.

Using Neuropixels 1.0 and 2.0 probes (Jun et al., 2017; Steinmetz et al., 2021), we recorded the activity of individual neurons in PFC, including anterior cingulate area (ACA), frontal pole (FRP), prelimbic area (PL), infralimbic area (ILA), and orbital area (ORB) (Laubach et al., 2018; Wang et al., 2020b). We also recorded from: secondary motor cortex (MOs), including anterolateral motor cotex (ALM), which has a well-characterized role in licking (Chen et al., 2017); olfactory cortex (OLF), including dorsal peduncular area (DP), dorsal taenia tecta (TTd), and anterior olfactory nucleus (AON), which receive input from the olfactory bulb (Igarashi et al., 2012; Mori and Sakano, 2021); and striatum, including caudoputamen (CP) and nucleus accumbens (ACB) (Fig. 1E-F), which are major outputs of PFC (Heilbronner et al., 2016). In a separate group of mice, we performed longitudinal 2-photon calcium imaging through a Gradient Refractive Index (GRIN) lens to track the activity of individual neurons in PL across several days of behavioral training (Fig. 1G-H). Both techniques permitted robust measurement of the activity of neurons of interest and generated complementary results (Figs. 1I-J, S2).

Graded cue and lick coding across the recorded regions

In the electrophysiology experiment, we isolated the spiking activity of 5332 individual neurons in regions of interest across 5 mice (449-1550 neurons per mouse, Figs. 2A, S3A). The activity of neurons in all regions exhibited varying degrees of modulation in response to the six cues (Fig. 2B). Broadly, there was strong modulation on CS+ and CS50 trials that appeared to be common to both odor sets (Fig. S3B). Across regions, there was heterogeneity in both the magnitude and the timing of the neural modulation relative to odor onset (Fig. S3C).

Graded cue and lick coding across the recorded regions.

(A) Location of each recorded neuron relative to bregma, projected onto 1 hemisphere. Each neuron is colored by CCF region. Numbers indicate total neurons passing quality control from each region.

(B) Mean normalized activity of all neurons from each region, aligned to odor onset, grouped by whether peak cue activity (0 - 2.5 s) was above (top) or below (bottom) baseline in held out trials. Number of neurons noted for each plot.

(C) Example kernel regression prediction of an individual neuron’s normalized activity on an example trial.

(D) CS+ trial activity from an example neuron and predictions with full model and with cues, licks, and reward removed. Numbers in parentheses are model performance (fraction of variance explained).

(E) Coordinates relative to bregma of every neuron encoding only cues or only licks, projected onto one hemisphere.

(F) Fraction of neurons in each region and region group classified as coding cues, licks, reward, or all combinations of the three.

(G) Additional cue (left) or lick (right) neurons in region on Y-axis compared to region on X-axis as a fraction of all neurons, for regions with statistically different proportions (see Methods).

To quantify the relative contribution of cues and conditioned responding (licking) to the activity of neurons in each region, we implemented reduced rank kernel regression (Steinmetz et al., 2019), using cues, licks, and rewards to predict neurons’ activity on held out trials (Figs. 2C, S5A). To determine the contribution of cues, licks, and rewards to each neuron’s activity, we calculated unique variance explained by individually removing each predictor from the model and calculating the reduction in model performance (Fig. 2D).

We identified individual neurons encoding cues, licks, or rewards as those for which that predictor uniquely contributed to 2% or more of their variance (a cutoff permitting no false positives and identifying neurons with robust task modulation, see Methods and Fig. S4). Neurons encoding cues (24% of all neurons), licks (11%), or both (16%) were most common. Neurons with any response to reward (independent of licking) were rare (5%) (Horst and Laubach, 2013). Cue neurons were characterized by sharp responses aligned to odor onset; in contrast, lick neurons’ responses were delayed and peaked around reward delivery (Fig. S5B-C), consistent with the timing of licks (Fig. 1C). The activity of cue neurons on rewarded and unrewarded CS50 trials validated our successful isolation of neurons with cue but not lick responses (Fig. S5D). The spatial distributions of cue and lick cells were noticeably different (Fig. 2E). The differences could be described as graded across regions, with the most lick neurons in ALM, and the most cue neurons in olfactory cortex and ORB, though each type of neuron was observed in every region (Fig. 2F-G, S6). Thus, our quantification of task encoding revealed varying proportions of cue and lick signaling across all regions.

Cue value coding is present in all regions

To expand upon our analysis identifying cue-responsive neurons, we next assessed the presence of cue value coding in this population. The three cue types (CS+, CS50, or CS-) in our behavioral task varied in relative value according to the predicted probability of reward (Eshel et al., 2016; Fiorillo et al., 2003; Winkelmeier et al., 2022). We reasoned that a neuron encoding cue value should have activity that scaled with the relative value of the cues (Fig. 3A). We modeled this relationship on a per-neuron basis by scaling a single cue kernel by its reward probability (0, 0.5, or 1, see Methods, Fig. 3B). This model describes cue activity as similar across odors of the same value, and scaling in magnitude according to each odor’s value. To consider alternative cue coding patterns, we also fit each neuron with 152 additional models containing all possible permutations of these values across the six cues, as well as models with selective responses for 1, 2, 3, 4, 5, or 6 cues, and determined which model best fit each neuron (Fig. S7). If cue responses were exclusively sensory and followed known olfactory coding properties (Pashkovski et al., 2020; Stettler and Axel, 2009), there would be no bias toward the ranked value model (CS+ > CS50 > CS-). We found, however, that this model was the most frequent best model, accounting for 14% of cue neurons (Fig. 3C). We refer to these neurons as value cells. There were two additional patterns that emerged across the population of cue neurons. First, there was a large fraction best explained by the model with equivalent responses to all 6 cues, which we term untuned cells (14% of cue neurons). Second, many of the alternative models had coding patterns that were similar to the ranked value model, and these appeared to be overrepresented among cue neurons, as well. We quantified the similarity to ranked value by correlating the values assigned to each cue in each model with those assigned to the cues in the ranked value model; this approach revealed an enrichment in neurons best fit by models most similar to ranked value (35% of cue neurons, Fig. 3C-D). We refer to neurons best fit by models most similar to the value model as value-like cells.

Robust value encoding and decoding among cue cells.

(A) Normalized activity of an example value cell with increasing modulation for cues with higher reward probability.

(B) For the same neuron, model-fit cue kernel for the original value model and with one of the 152 alternatively-permuted cue coding models.

(C) Distribution of best model fits across all cue neurons. Light blue is value model, purple is value-like models, gray is untuned model, and the remaining models are dark blue. Value-like models are shaded according to their correlation with ranked value, as illustrated in (D). Dashed line is chance proportion when assuming even distribution.

(D) Schematic of value assigned to each of the 6 cues for many of the cue coding models (full schematic in Fig. S7). Value-like models are sorted by their correlation with the ranked value model.

(E) Left: normalized activity of every value cell, sorted by mean firing 0 - 1.5 s following odor set A CS+ onset. Right: mean normalized activity of all value cells, grouped by whether peak cue activity (0 - 2.5 s) was above (top) or below (bottom) baseline in held out trials. Number of neurons noted for each plot.

(F) As in (E), for value-like cells.

(G) Accuracy (mean +/− SEM across neurons) of decoded cue identity for single neurons of value, value-like, and untuned neurons. * indicates where value, value-like, and untuned neurons significantly differed from each other and baseline (all p < 0.001, Bonferroni corrected). All pairwise comparisons in Supplemental Table 2.

(H) Accuracy (mean+/−std across bootstrapped iterations) of decoded cue identity using different numbers of neurons.

(I) Left: estimated value (mean +/− SD across 1000 bootstrapped iterations) of held out CS+ (top) and CS- (bottom) trials using linear models trained on the activity of value, value-like, or untuned neurons. Right: accuracy (mean+/−std across bootstrapped iterations) of decoded cue value using these value estimates. * indicates where the accuracy of value neurons exceeded value-like and untuned neurons (all p < 0.016, bootstrapped). All pairwise comparisons in Supplemental Table 2.

Interestingly, the frequency of value cells was similar across the recorded regions (Fig. 4A). Despite the regional variability in number of cue cells broadly (Fig. 2F-G), there were very few regions that statistically differed in their proportions of value cells (Figs. 4A, S9). Overall, there were slightly more value cells across all of PFC than in motor and olfactory cortex (Figs. 4A, S9). Although olfactory cortex had the most cue cells, these were less likely to encode value than cue cells in other regions (Fig. S10). Value-like cells were also widespread; they were less frequent in motor cortex as a fraction of all neurons, but they were equivalently distributed in all regions as a fraction of cue neurons (Figs. 4B, S9, S10).

Widespread cue value coding.

(A) Fraction of neurons in each region and region group classified as value cells (blue) and other cue neurons (gray), as well as fraction (+/− 95% CI) estimated from a linear mixed effects model with random effect of session (see Methods). PFC has more value cells than motor (p = 0.002) and olfactory (p = 0.00005) cortex. All pairwise comparisons in Supplemental Table 3.

(B) As in (A), for value-like cells. Motor cortex has fewer value-like cells than PFC (p = 8 10−6) and olfactory cortex (p = 4 10−8). All pairwise comparisons in Supplemental Table 3.

(C) First principal component value cells from all regions.

(D) As in (C), for value-like cells.

(E) Accuracy of decoded cue value (mean +/− SD across 1000 bootstrapped iterations) as in Fig. 3I, using 5 (with replacement) value cells from each region (left) and 25 value cells from each region group (right) using cue-evoked (blue) and baseline (black) activity. No regions or region groups significantly differed from each other (p > 0.46, Bonferroni corrected). All pairwise comparisons in Supplemental Table 3.

We next investigated the robustness of the value representation in each of our recorded regions. Principal component analysis on value and value-like cells from each region revealed similarly strong value-related dynamics across motor, prefrontal, and olfactory regions (Fig. 4C-D). We quantified the robustness of value coding in each region by decoding cue value using selections of value cells from each region and found similar performance across all regions (Fig. 9E). Taken together, these data illustrate that, in contrast to cue and lick coding broadly, value coding is similarly represented across the regions we sampled. In fact, this observation extended to the striatal regions we sampled as well, indicating that such value coding is widespread even beyond cortex (Fig. S11).

Because cue valuations can be influenced by preceding reward outcomes, we next considered whether the cue value signaling we detected was sensitive to the history of reinforcement (Nakahara et al., 2004; Ottenheimer et al., 2020; Winkelmeier et al., 2022). To estimate the subjects’ trial-by-trial cue valuation, we fit a linear model predicting the number of anticipatory licks on each trial using cue type, overall reward history, and cue type-specific reward history as predictors. We found a strong influence of cue type-specific reward history and a more modest influence of overall reward history (Fig. 5A). We used the model prediction of licks per trial as our estimate of trial value; the effects of reward history on lick rate were apparent when grouping trials by the value estimates from the trial value model (Fig. 5B).

A subset of cue cells incorporate reward history.

(A) Coefficient weight (+/− standard error from model fit) for reward outcome on the previous 10 trials of any type (left) and on the previous 10 trials of the same cue type (right) for the ‘trial value’ model: a linear model predicting the number of anticipatory licks on every trial of every session. Lick rates were normalized so that the maximum lick rate for each session was equal to 1. Colored lines are models fit to each individual mouse.

(B) Mean (+/− SEM) lick rate across mice (n = 5 mice) on trials binned according to value estimated from the trial value model.

(C) Normalized activity of an example history value cell with increasing modulation for cues of higher value.

(D) For the same neuron, model-predicted activity with the original value model (left) and with the history model, which uses trial-by-trial value estimates from the trial value model (right).

(E) For the same neuron, model-predicted activity using licks. Inset: variance explained using licks versus history for history neurons.

(F) The activity of all cells in each category projected onto the coding dimension maximally separating CS- and CS+ for trials binned by value estimated from the trial value model.

(G) The mean (+/− std across 5000 bootstrapped selections of neurons) activity (1 - 2.5 s from odor onset) along the coding dimension maximally separating CS- and CS+ for trials binned by value estimated from the lick model.

(H) The mean (+/− std across 5000 bootstrapped selections of neurons) slope of the activity on CS50 trials regressed onto the trial value model estimate for those trials. History and lick cells had greater slopes than the other groups (p < 0.0003, see Supplemental Table 4).

(I) Fraction of neurons in each region and region group classified as history cells (light blue) and other cue neurons (gray), as well as estimated fraction (+/−95% CI) with random effect of session (see Methods). PFC had more history cells than motor (p = 0.0016) and olfactory (p = 0.00053) cortex. All pairwise comparisons in Supplemental Table 4.

We therefore investigated whether value cells showed similar trial-by-trial differences in their cue-evoked firing rates (Fig. 5C). To test this, we compared the fit of our original cue coding models (Fig. 3B-D) with an alternative model in which the kernel scaled with the pertrial value estimates from our trial value model (Fig. 5D). Overall, 5% of cue cells, including 15% of value cells, were best fit by the history model. Although the number of anticipatory licks per trial was used to generate the trial value estimates, the precise licking pattern on those trials was a poorer predictor of neural responses than the trial value-scaled cue kernel model (Fig. 5E). To further evaluate the history component of these neurons, we calculated these neurons’ activity on CS50 trials of varying value estimates from the trial value model and projected it onto the population dimension maximizing the separation between CS+ and CS-. We hypothesized that high value CS50 trials would be closer to CS+ activity while low value CS50 trials would be closer to CS-activity. Indeed, history cells (and lick cells) demonstrated graded activity along this dimension, in contrast to non-history value, value-like, and untuned cells (Fig. 5F-H). Finally, we examined the regional distribution of history cells and found low numbers across all regions, but with higher prevalence overall in PFC than in motor and olfactory cortex (Fig. 5I), lending additional support for slightly enhanced value coding in PFC.

Cue coding emerges along with behavioral learning

To determine the timescales over which these coding schemes emerged and persisted, we performed longitudinal 2-photon calcium imaging and tracked the activity of individual neurons across several days of behavioral training (Fig. 6A). We targeted a GRIN lens to PL, a location with robust cue and lick coding (Fig. 2F) and where cue responses were predominantly value or value-like (Figs. 3E-F, S10). Mice (n = 8) developed anticipatory licking during the first sessions of odor set A (A1) that differentiated CS+ trials from CS50 (t(7) = 3.2, p = 0.015) and CS-(t(7) = 7.0, p = 0.0002) trials and CS50 trials from CS- (t(7) = 3.7, p = 0.008) trials (Fig. 6B-C). Visualizing the normalized activity across the imaging plane following CS+ presentation early and late in session A1 revealed a pronounced increase in modulation across this first session (Fig. 6D-E). Individual neurons (n = 705, 41-165 per mouse) also displayed a notable increase in modulation in response to the CS+ after task learning (Fig. 6F).

Acquisition of conditioned behavior and cue encoding in PFC.

(A) Training schedule for 5 of the mice in the calcium imaging experiment. An additional 3 were trained only on odor set A.

(B) Mean (+/− SEM) licking on early (first 60) and late (last 60) trials from day 1 of odor set A (n = 8 mice).

(C) Mean (+/− SEM) baseline-subtracted anticipatory licks for early and late trials from each day of odor set A. Thin lines are individual mice (n = 8 mice).

(D) Standard deviation of fluorescence from example imaging plane.

(E) Normalized activity of each pixel following CS+ presentation on early and late trials of session A1.

(F) Normalized deconvolved spike rate of all individual neurons on early and late trials of session A1.

(G) Proportion of neurons classified as coding cues, licks, rewards, and all combinations for each third of session A1.

(H) Mean(+/− SEM) unique variance explained by cues, licks, and rewards for neurons from each mouse. Thin lines are individual mice. Unique variance was significantly different across session thirds for cues (F (2, 21) = 3.71, p = 0.04) but not licks (F (2, 21) = 0.37, p = 0.69) or reward (F (2, 21) = 0.65, p = 0.53, n = 8 mice, one-way ANOVA).

(I) Mean (+/− SEM) normalized deconvolved spike rate for cells coding cues, licks, both, or neither on early and late trials, sorted by whether peak cue activity (0 - 2.5 s) was above (top) or below (bottom) baseline for late trials.

To determine whether this increase in activity was best explained by a cue-evoked response, licking, or both, we again used kernel regression to fit and predict the activity of each neuron for early, middle, and late trials in session A1. The number of individual neurons encoding cues more than doubled from early to late A1 trials (Fig. 6G). The unique variance cues increased across this first session, in contrast to licks and reward (Fig. 6H). This stark change in cue coding was also noticeable when plotting neurons encoding cues, licks, or both, as defined at the end of the session, on both early and late trials (Fig. 6I). These data indicated that PFC neural activity related to cues (but not licks) rapidly emerges during initial learning of the behavioral task.

Cue and lick coding is stable across days

We next assessed whether cue and lick coding were stable across days. By revisiting the same imaging plane on each day of training, we were able to identify neurons that were present on all three days of odor set A training (n = 371, 20-65 per mouse) (Fig. 7A-B). There was remarkable conservation of task responding across days, both on an individual neuron level (Fig. 7C) and across all imaged neurons (Fig. 7D). In fact, neurons were much more correlated with their own activity on the subsequent day than would be expected by chance (Figs. 7E, S12A). To further quantify coding stability, we fit our kernel regression to the activity of each neuron on session A3 (Fig. 7F) and then used these models to predict activity on early, middle, and late trials on sessions A1-3. Session A3 model predictions were most highly correlated with true activity during A3, but they outperformed shuffle controls at all time points, demonstrating preservation of a learned coding scheme (Figs. 7G, S12B). We then asked more specifically whether cells coding cues, licks, and both maintained their coding preferences across days. For each group of cells, we calculated their unique cue, lick, and reward variance at each time point. The preferred coding of each group, as defined on session A3, was preserved in earlier days (Fig. 7H). Thus, cue and lick coding are stable properties of PFC neurons across multiple days of behavioral training.

Cue and lick coding is stable across days.

(A) Standard deviation fluorescence from example imaging plane.

(B) Masks (randomly colored) for all tracked neurons from this imaging plane.

(C) Deconvolved spike rate on every CS+ trial from all three sessions of odor set A for an example neuron. Vertical dashed line is reward delivery. Color axis as in (D).

(D) Normalized deconvolved spike rate for all tracked neurons on all three sessions of odor set A.

(E) Correlation between the activity of a given neuron in one session and its own activity in the subsequent session, quantified as a percentile out of correlations with the activity of all other neurons on the subsequent day. Plotted as the median for each subject and the mean (+/− SEM) across these values. Real data was more correlated than shuffled data (p = 0.0078 for both comparisons, Wilcoxon signed-rank test).

(F) Fraction of tracked neurons coding cues, licks, rewards, and their combinations on day 3.

(G) Model performance when using models from session A3 to predict the activity of individual neurons across session thirds of odor set A training, plotted as mean (+/− SEM) correlation between true and predicted activity across mice, normalized to the correlation between model and training data. Thin lines are individual mice. Performance was greater than shuffled data at all time points (p < 0.002, Bonferroni-corrected, n = 8 mice). Non-normalized data in Fig. S12.

(H) Mean (+/− SEM) unique cue, lick, and reward variance for cells classified as coding cues, licks, both, or neither on session A3. A3 cue cells had increased cue variance in A2 (p < 10−7, see Methods) and A1 (p < 0.03) relative to lick and reward variance. Same pattern for A3 lick cells in A2 (p < 0.0001) and A1 (p < 0.01).

A subset of mice (n = 5) also learned a second odor set (odor set B), presented on separate days. Activity was very similar for both odor sets, evident across the entire imaging plane (Fig. 8A), for individual tracked neurons (n = 594, 81-153 per mouse) (Fig. S2B), and for kernel regression classification of these neurons (Fig. 8B). Notably, odor set A models performed similarly well at predicting both odor set A and odor set B activity (Fig. 8C). Moreover, cue, lick and both neurons maintained their unique variance preference across odor sets (Fig. 8D). Finally, to investigate the presence of value coding across odor sets over separate days, we fit tracked cue neurons with the value model and its shuffles. Even with odor sets imaged on separate days (days 5 and 6 of training, A3 and B3), we again found that the value and value-like models were the best models for sizable fractions (9% and 47%, respectively) of cue neurons, demonstrating that value coding is conserved across stimulus sets on consecutive days (Fig. 8E-G). Given the prominence of value-like signals in this imaged population, we then assessed the stability of cue cells with preferential CS+ responses across the tracked A1-3 sessions and found conservation of a value-like coding pattern (Fig. 8H) and, as with the whole population (Fig. 7G), greater correlation in activity across days than expected by chance (Fig. 8I).

Cue and lick coding in separately trained odor sets.

(A) Normalized activity of all pixels in the imaging plane following CS+ presentation on the third day of each odor set (A3 and B3, days 5 and 6 of training).

(B) Fraction of neurons coding for cues, licks, rewards, and their combinations in A3 and B3 (days 5 and 6).

(C) Mean (+/− SEM, across mice) correlation between activity predicted by odor set A3 models and its training data (A3, cross-validated) or activity in B3, for true (black) and trial shuffled (gray) activity. Thin lines are individual mice. F (1, 16) = 3.2, p = 0.09 for main effect of odor set, F (1, 16) = 135, p < 10−8 for main effect of shuffle, F (1, 16) = 2.2, p = 0.16 for interaction, n = 5 mice, two-way ANOVA.

(D) Mean (+/− SEM, across mice) unique cue, lick, and reward variance for cells classified as coding cues, licks, both, or neither for odor set A. For each category, odor set A unique variance preference was maintained for odor set B (p < 0.04) except for both cells, for which lick and reward variance were not different in odor set B (p = 0.22, Bonferroni-corrected, n = 5 mice).

(E) Distribution of best model fits across all cue cells, with colors from Fig. 3C. Dashed line is chance proportion when assuming even distribution.

(F) Left: normalized activity of every value cell, sorted by mean firing 0 - 1.5 s following odor set A CS+ onset. Right: mean normalized activity of all value cells, grouped by whether peak cue activity (0 - 2.5 s) was above (top) or below (bottom) baseline in held out trials. Number of neurons noted for each plot.

(G) As in (E), for value-like cells.

(H) Mean (+/− SEM, across neurons) activity of cue cells tracked across A1, A2, and A3 with preferential CS+ firing, defined on half of A3 trials and plotted for the other half of A3 trials and all of A1 and A2 trials.

(I) For neurons in (H), correlation between a neuron’s activity in one session and its own activity in the subsequent session, quantified as a percentile out of correlations with the activity of all other neurons on the subsequent day. Plotted as the median for each subject (n = 7 with CS+ preferring cue cells) and the mean (+/− SEM) across these values. Real data was more correlated than shuffled data (p = 0.016 A1:A2, p = 0.031 A2:A3, Wilcoxon signed-rank test).

Discussion

Our experiments assessed how coding for reward-predicting cues and reward-seeking actions differed across brain regions and across multiple days of training. We found coding for cues and licks in all regions we sampled, but their proportions varied in a graded way across those regions. In contrast to regional differences in the proportion of cue-responsive neurons, cue value cells were present in all regions and value could be decoded from them with similar accuracy regardless of region. Coding for cue value was greatly overrepresented compared to alternative cue coding schemes and, in a subset of neurons, incorporated the recent reward history. Cue coding was established within the first day of training and neurons encoding cues or licks maintained their coding preference across multiple days of the task; the value characteristics of cue cells were also maintained across days. These results demonstrate widespread value coding and stability of cue and lick codes in PFC.

Graded cue and lick coding across regions

We found robust and separable coding for licks and cues (and combined coding of both) in all regions using electrophysiology and in PL using calcium imaging. The widespread presence of lick coding is consistent with recent reports of distributed movement and action coding (Musall et al., 2019; Steinmetz et al., 2019; Stringer et al., 2019); however, we saw sizable differences in the amount of lick coding across recorded regions. Notably, ALM had the greatest number of lick neurons, as well as the fewest cue neurons, perhaps reflecting its specialized role in the preparation and execution of licking behavior (Chen et al., 2017). Conversely, the olfactory cortical regions DP, TTd, and AON had the most cue neurons (especially non-value coding cue neurons), suggesting a role in early odor identification and processing (Mori and Sakano, 2021). PFC subregions balanced lick and cue coding, consistent with their proposed roles as association areas (Klein-Flügge et al., 2022; Miller and Cohen, 2001), but there was variability within PFC as well. In particular, ORB had a greater fraction of cue cells than any other subregions, consistent with its known dense inputs from the olfactory system (Ekstrand et al., 2001; Price, 1985; Price et al., 1991). Thus, our results establish that the neural correlates of this Pavlovian conditioned behavior consist of a gradient of cue and response coding rather than segmentation of sensory and motor responses.

Widespread value signaling

Value signals can take on many forms and occur throughout task epochs. In our experiments, we focused on the predicted value associated with each conditioned stimulus, which is crucial for understanding how predictive stimuli produce motivated behavior (Berridge, 2004). Surveys of value coding in primate PFC have found individual neurons correlated with stimulus-predicted value in many subregions, with the strongest representations typically in ORB (Hunt et al., 2018; Kennerley et al., 2009; Roesch and Olson, 2004; Sallet et al., 2007). In rodents, there is also a rich literature on value signaling in ORB (Kuwabara et al., 2020; Namboodiri et al., 2019; Schoenbaum et al., 2003; Stalnaker et al., 2014; Sul et al., 2010; van Duuren et al., 2009; Wang et al., 2020a), but there have also been many reports of value-like signals in frontal cortical regions beyond ORB (Allen et al., 2019; Kondo and Matsuzaki, 2021; Otis et al., 2017; Wang et al., 2020a). In our present experiment, we sought to expand upon these rodent results by separating cue activity from licking, which tracks value and may confound interpretation, by including more than two cue types, which provided a rich space to assess value coding, and by sampling from many frontal regions in the same experiment.

When considering the number of neurons responsive to cues rather than licks, our data confirmed the importance of ORB, which has more cue-responsive neurons than motor and other prefrontal regions, but, beyond cue responsivity, we were interested in identifying specific cue coding patterns pertaining to value. By analyzing the activity of cue-responsive neurons across all 6 odors predicting varying probabilities of reward, we were able to isolate neurons coding value, as well as those with value-like signals that could easily be misconstrued as value-coding in a task with fewer cues and value levels. Included in the value-like models are coding patterns that bias their activity for higher value odors without fitting our strict linear ranked value criteria; for instance, selective firing for one or two of the CS+ odors. The enrichment of these models among cue responsive neurons, even in olfactory cortex, indicates the prevalence of value-biased coding schemes for odor-responsive neurons across brain regions. The question remains of where odor information is first shaped according to value. There have been multiple reports of some association-related modification of odor representations as early as the olfactory bulb (Chu et al., 2016; Doucette et al., 2011; Koldaeva et al., 2019; Li et al., 2015). Considering we detected value and especially value-like coding in AON, DP, and TTd, perhaps these regions are a crucial first step in processing and amplifying task-related input from the olfactory bulb. Because they provide input to PFC (Bhattarai et al., 2021; Igarashi et al., 2012), they may be an important source of the cue coding we observed there.

The distribution of cue cells with linear coding of value was mostly even across regions, with slight enrichment overall in PFC compared to motor and olfactory cortex, but no subregional differences in PFC. Importantly, cue value could be decoded from value cells in each region with similar accuracy. One consequence of a widely distributed value signal is that manipulating only one subregion would be less likely to fully disrupt value representations, which is consistent with the results of studies comparing functional manipulations across PFC (Chudasama and Robbins, 2003; Dalton et al., 2016; St. Onge and Floresco, 2010; Verharen et al., 2020; Wang et al., 2020a). Different subregional impacts on behavior may reveal biases in how the value signal in each region contributes to reward-related behaviors, for instance during learning or expression of a reward association (Namboodiri et al., 2019; Otis et al., 2017; Wang et al., 2020a). A related interpretation is that, in this task, there may be other properties that correlate with cue value, and the homogeneous value representation we observed across regions masks regional differences in tuning to these other correlated features, such as motivation (Roesch and Olson, 2004) and a host of related concepts, including salience, uncertainty, vigor, and arousal (Hayden and Niv, 2021; Stalnaker et al., 2015; Zhou et al., 2021), which can have different contributions to behavior. This interpretation is consistent with broader views that observations of ‘value’ signals are often misconstrued (Zhou et al., 2021) and that pure abstract value may not be encoded in the brain at all (Hayden and Niv, 2021). Although the identification of value in our task was robust to three levels of reward probability across two stimulus sets, the fact that this signal was widespread contributes to the case for revisiting the definition and interpretation of value to better understand regional specialization.

In our analysis, we uncovered a distinction between neurons encoding the overall value of cues and those with value representations that incorporated the recent reward history. Neurons with history effects were rare and most frequent in PFC. These neurons may have a more direct impact on behavioral output in this task, because the lick rate also incorporated recent reward history. Notably, the impact of reward history on these neurons was noticeable even prior to cue onset, consistent with a previously proposed mechanism for persistent value representations encoded in the baseline firing rates of PFC neurons (Bari et al., 2019).

Stability of PFC codes

Previous reports have observed drifting representations in PFC across time (Hyman et al., 2012; Malagon-Vina et al., 2018), and there is compelling evidence that odor representations in piriform drift over weeks when odors are experienced infrequently (Schoonover et al., 2021). On the other hand, it has been shown that coding for odor association is stable in ORB and PL, and that coding for odor identity is stable in piriform (Wang et al., 2020a), with similar findings for auditory Pavlovian cue encoding in PL (Grant et al., 2021; Otis et al., 2017) and ORB (Namboodiri et al., 2019). We were able to expand upon these data in PL by identifying both cue and lick coding and showing separable, stable coding of cues and licks across days and across sets of odors trained on separate days. We were also able to detect value coding common to two stimulus sets presented on separate days, and conserved value features across the three training sessions. Notably, the model with responses only to CS+ cues best fit a larger fraction of imaged PL neurons than the ranked value model, a departure from the electrophysiology results. It would be interesting to know if this is due to a bias introduced by the calcium imaging approach, the slightly reduced CS50 licking relative to CS+ licking in the imaging cohort, or the shorter imaging experimental timeline.

The consistency in cue and lick representations we observed indicates that PL serves as a reliable source of information about cue associations and licking during reward seeking tasks, perhaps contrasting with other representations in PFC (Hyman et al., 2012; Malagon-Vina et al., 2018). Interestingly, the presence of lick, but not cue coding at the very beginning of the first session of training suggests that lick cells in PL are not specific to the task but that cue cells are specific to the learned cue-reward associations. Future work could expand upon these findings by examining stimulus-independent within session value coding across many consecutive days.

Overall, our work emphasizes the importance of evaluating regional specialization of neural encoding with systematic recordings in many regions using the same task. Future work will clarify whether cue value is similarly widely represented in other reward-seeking settings and whether there are regional differences in the function of the value signal.

Supporting information